Escape Analysis and Specialization in a Tracing JIT

### 2010-09-27 talk/pepm2011/paper.tex
More tweaks.

The goal of a just-in-time compiler for a dynamic language is obviously to
improve the speed of the language over an implementation of the language that
uses interpretation. The first goal of a JIT is thus to remove the
interpretation overhead, i.e. the overhead of bytecode (or AST) dispatch and the
overhead of the interpreter's data structures, such as operand stack etc. The
second important problem that any JIT for a dynamic language needs to solve is
how to deal with the overhead of boxing of primitive types and of type
dispatching. Those are problems that are usually not present or at least less
severe in statically typed languages.

@@

A recently popular approach to implementing just-in-time compilers for dynamic
languages is that of a tracing JIT. A tracing JIT often takes the form of an
extension to an existing interpreter, which can be sped up that way. This
approach is also the one taken by the PyPy project, which is an environment for
implementing dynamic programming languages. PyPy's
approach to doing so is to straightforwardly implement an interpreter for the
to-be-implemented language, and then use powerful tools to turn the interpreter
into an efficient VM that also contains a just-in-time compiler. This compiler
is automatically generated from the interpreter using partial-evaluation-like
techniques \cite{bolz_tracing_2009}. The PyPy project and its approach to
tracing JIT compilers is described in Section~\ref{sec:Background}.

@@

The work described in this paper was done in the context of the PyPy project
\cite{armin_rigo_pypys_2006}. PyPy is an environment where dynamic languages can
be implemented in a simple yet efficient way. The approach taken when
implementing a language with PyPy is to write an interpreter for the language in
\emph{RPython} \cite{davide_ancona_rpython:_2007}. RPython ("restricted Python")
is a subset of Python chosen in such a way, that type inference becomes
possible. The language interpreter can thus be translated with the help of
PyPy's tools into a VM on the C level. Because the interpreter is written at a
relatively high level, the language implementation is kept free of low-level
details, such as object layout, garbage collection or memory model. Those
aspects of the final VM are woven into the generated code during the translation
to C. XXX languages that are done using PyPy

The feature that makes PyPy more than a compiler with a runtime system is it's
support for automated JIT compiler generation \cite{bolz_tracing_2009}. During
the translation to C, PyPy's tools can generate a just-in-time compiler for the
language that the interpreter is implementing. This process is not fully
automatic, but needs to be guided by the language implementer by some
source-code hints. Semi-automatically generating a JIT compiler has many advantages
over writing one manually, which is an error-prone and tedious process. The
generated JIT has the same semantics as the interpreter by construction, and all
languages implemented using PyPy benefit from improvements to the JIT generator.
The JIT that is produced by PyPy's JIT generator is a \emph{tracing JIT
compiler}, a concept which will be explained in more detail in the next section.

@@

Tracing JITs are a recently more popular approach to write just-in-time
compilers for dynamic languages \cite{XXX}. Their origins lie in the Dynamo
project, that used a tracing approach to optimize machine code using execution
traces \cite{XXX}. They were then adapted to be used for a very light-weight
Java VM \cite{XXX} and afterwards used in several implementations of dynamic
languages, such as JavaScript \cite{XXX}, Lua \cite{XXX} and now Python via
PyPy.

@@

\begin{figure}
\includegraphics[scale=0.7]{figures/obj-lifetime.pdf}

>>>>

The goal of a just-in-time (JIT) compiler for a dynamic language is obviously to
improve the speed of the language over an implementation of the language that
uses interpretation. The first goal of a JIT is thus to remove the
interpretation overhead, i.e. the overhead of bytecode (or AST) dispatch and the
overhead of the interpreter's data structures, such as operand stack etc. The
second important problem that any JIT for a dynamic language needs to solve is
how to deal with the overhead of boxing of primitive types and of type
dispatching. Those are problems that are usually not present or at least less
severe in statically typed languages.

@@

A recently popular approach to implementing just-in-time compilers for dynamic
languages is that of a tracing JIT. A tracing JIT often takes the form of an
extension to an existing interpreter, which can be sped up that way. This
approach is also the one taken by the PyPy project, which is an environment for
implementing dynamic programming languages. PyPy's
approach to doing so is to straightforwardly implement an interpreter for the
to-be-implemented language, and then use powerful tools to turn the interpreter
into an efficient virtual machine (VM) that also contains a just-in-time compiler.
This compiler
is automatically generated from the interpreter using partial-evaluation-like
techniques \cite{bolz_tracing_2009}. The PyPy project and its approach to
tracing JIT compilers is described in Section~\ref{sec:Background}.

@@

The work described in this paper was done in the context of the PyPy project
\cite{armin_rigo_pypys_2006}. PyPy is an environment where dynamic languages can
be implemented in a simple yet efficient way. The approach taken when
implementing a language with PyPy is to write an \emph{interpreter}
for the language in
\emph{RPython} \cite{davide_ancona_rpython:_2007}. RPython ("restricted Python")
is a subset of Python chosen in such a way that type inference becomes
possible. The language interpreter can then be compiled (``translated'') with
PyPy's tools into a VM on the C level. Because the interpreter is written at a
relatively high level, the language implementation is kept free of low-level
details, such as object layout, garbage collection or memory model. Those
aspects of the final VM are woven into the generated code during the translation
to C. XXX languages that are done using PyPy

The feature that makes PyPy more than a compiler with a runtime system is it's
support for automated JIT compiler generation \cite{bolz_tracing_2009}. During
the translation to C, PyPy's tools can generate a just-in-time compiler for the
language that the interpreter is implementing. This process is mostly
automatic; it only needs to be guided by the language implementer by a small number of
source-code hints. Mostly-automatically generating a JIT compiler has many advantages
over writing one manually, which is an error-prone and tedious process.
By construction, the generated JIT has the same semantics as the interpreter, and
the process benefits all languages implemented as an interpreter in RPython.
The JIT that is produced by PyPy's JIT generator is a \emph{tracing JIT
compiler}, a concept which we now explain in more details.

@@

Tracing JITs are a recently popular approach to write just-in-time
compilers for dynamic languages \cite{XXX}. Their origins lie in the Dynamo
project, which used a tracing approach to optimize machine code using execution
traces \cite{XXX}. Tracing JITs have then be adapted to be used for a very light-weight
Java VM \cite{XXX} and afterwards used in several implementations of dynamic
languages, such as JavaScript \cite{XXX}, Lua \cite{XXX} and now Python via
PyPy.

@@

\begin{figure}
\begin{center}
\includegraphics[scale=0.7]{figures/obj-lifetime.pdf}
\end{center}
